Preferences for the commodification of pensions in Chile: the rol of intergenerational social mobility

### 2025-11-30 paper/02-antecedents.qmd
Up antecedentes

@@

## Preferences for market-based welfare

Beyond state’s capacity to reallocate resources from the advantaged to the vulnerable, market institutions also play a central role in distributing socially valuable goods and rewards [@koos_moral_2019; @lindh_bringing_2023]. As Polanyi [-@polanyi_great_1975] observed, economic integration in capitalist societies is primarily organized through market exchange, governed by a self-regulating price system embedded in institutional frameworks. These institutions are not mere aggregates of individual behavior, but social realities endowed with rules, mechanisms, and normative meanings that shape everyday thinking [@koos_moral_2019; @immergut_theoretical_1998]. In this sense, the economic order is mirrored in a moral economy: collectively shared norms and beliefs about justice in distribution, embedded and reinforced through institutions [@mau_inequality_2015; @svallfors_political_2007]. While most research from this perspective has focused on welfare institutions, recent scholarship has brought the market back into focus as a site of distributive justice beliefs and institutional responses to inequality [@lindh_bringing_2023]. In many countries, privatization and commodification have expanded market logic into core areas of social reproduction, such as healthcare, education, pensions, and social security, deepening inequality in access to these services [@gingrich_making_2011; @ferre_welfare_2023]. Yet public support for market-based welfare provision has grown, even in traditional welfare states, particularly among higher-income groups who view private alternatives as more efficient or higher in quality [@busemeyer_welfare_2020]. This shift calls for broader inquiry into how institutions like markets structure access to resources and legitimize inequalities [@satz_por_2019; @mau_inequality_2015].

The legitimacy of market-based inequalities is closely tied to beliefs about distributive justice grounded in market principles. Beyond stratification structures and socioeconomic conditions, the study of inequality also focuses on individuals’ beliefs about the origins of inequality, the normative frameworks that sustain those beliefs, the mechanisms that shape them, and their implications for attitudes and behavior [@kluegel_beliefs_1981]. In this regard, empirical research on distributive justice focuses on individuals’ conceptions of how goods and rewards _should_ be distributed in society [@jasso_distributive_2016]. This line of inquiry allows for examining the extent to which economic inequality is perceived as just, or in a certain sense, legitimate [@castillo_legitimacy_2011]. Most of the research on this field explores the legitimacy of wage inequality, specially salary gaps between jobs at opposite ends of the occupational hierarchy [@castillo_legitimacy_2011; @jasso_gender_1999; @jasso_justice_1978]. Recently, others examines how individuals justify market-generated inequalities in access to core social services such as healthcare, education, pensions or social security. Here, legitimacy stems from the belief that access to these goods should follow market-based criteria [@castillo_socialization_2024;  @lindh_public_2015]. In such views, these services are treated as legitimate commodities; goods that can be traded, priced, and evaluated through market logic [@busemeyer_welfare_2020].

Market justice preferences refer to normative beliefs that legitimize the idea that access to core social services—such as healthcare, education, or pensions—should be determined by market-based criteria. Following Janmaat’s [-@janmaat_subjective_2013, p. 359] distinction, these preferences fall under the category of “beliefs,” understood as normative ideas about what inequality should look like, as opposed to “perceptions,” which refer to subjective evaluations of existing inequality. Market justice preferences reflect the view that access to these services should depend on individuals’ ability to pay, thus justifying inequalities generated by market mechanisms [@kluegel_legitimation_1999;  @lindh_public_2015]. The concept draws on Lane’s [-@lane_market_1986] classic contrast between market justice and political justice: the former is grounded in the idea of earned deserts—where rewards reflect effort, productivity, and skill—while the latter prioritizes need and equality, typically expressed in welfare state policies. Lane argued that markets and states differ in purpose (efficiency vs. need), logic (individual vs. collective), and fairness criteria (merit vs. equality). Market justice assumes that markets are neutral, self-regulating systems in which fair procedures yield outcomes proportional to merit. Inequality, from this perspective, is not only expected but legitimate—as long as it arises from fair competition. In this way, market justice offers a moral lens through which individuals can view the commodification and stratified access to social services as fair and justified [@kluegel_legitimation_1999; @lindh_public_2015].

The empirical study of market justice preferences has employed various strategies to capture how individuals assess inequalities arising from market allocation. A common approach evaluates whether people find it fair that access to essential social services—such as healthcare, education, or pensions—depends on income. This builds on foundational work by Kluegel and Smith [-@kluegel_legitimation_1999], who examined normative justifications for capitalist systems. Subsequent studies have extended this logic to a wider set of welfare goods. For example, Immergut and Schneider [-@immergut_it_2020] and von dem Knesebeck et al. [-@vondemknesebeck_are_2016] explore whether respondents believe it is just that higher-income individuals receive better healthcare. Similarly, Lee and Stacey [-@lee_fairness_2023] and @castillo_changes_2025 apply this framework to education. Complementing these efforts, more recent research has introduced composite indicators to capture broader orientations toward market-based allocation. Lindh [-@lindh_public_2015], for instance, constructs an index averaging support for income-based access to healthcare and education in comparative perspective, while Castillo et al. [-@castillo_socialization_2024] propose a single-item measure covering health, education, and pensions in Chile. These instruments seek to gauge to what extent individuals view market-generated inequalities as legitimate. Taken together, they capture two core dimensions of market distribution: the role of economic resources as a key determinant of outcomes, and the framing of services as tradable commodities that can be bought and sold according to ability to pay [@lindh_public_2015].

Comparative empirical research has identified several individual-level factors that influence support for market justice. Individuals in more advantaged socioeconomic positions—those with higher income, education, and occupational status—are consistently more likely to endorse market-based distributive principles [@koos_moral_2019; @lindh_public_2015; @svallfors_political_2007]. For example, Lindh [-@lindh_public_2015] finds that individuals from the service class are more likely to support market-based access to healthcare and education than skilled and unskilled workers across 17 relatively affluent countries. In a comparative analysis, Svallfors [-@svallfors_political_2007] observes that this expected class pattern appears clearly only in Sweden, where support for private education and healthcare varies systematically by class. Busemeyer [-@busemeyer_skills_2014] similarly shows that support for private education is stronger among high-income groups, while Immergut and Schneider [-@immergut_it_2020] and von dem Knesebeck et al. [-@vondemknesebeck_are_2016] report comparable findings for healthcare, suggesting that wealthier individuals perceive private provision as a means of maintaining relative advantage. In Chile, Otero and Mendoza [-@otero_power_2024] show that individuals with higher income and university education express stronger support for market allocation in healthcare, education, and pensions. Beliefs about inequality and political orientation also matter. Castillo et al. [-@castillo_socialization_2024] and @castillo_perceptions_2025 show that individuals who have strong meritocratic beliefs are more likely to support market-based distribution in Chile, while Lee and Stacy [-@lee_fairness_2023] in Australia suggest that these preferences are also greater as people lean toward economic conservatism. In this sense, market justice preferences are shaped by the interplay between structural position and normative reasoning.

However, individual characteristics alone do not fully explain variation in support for market justice. A growing body of cross-national research highlights the importance of institutional arrangements in shaping these preferences. For instance, Immergut and Schneider [-@immergut_it_2020] find that in countries with higher public spending on healthcare, individuals are less likely to view income-based access as fair. Similarly, Busemeyer [-@busemeyer_skills_2014] shows that increased public investment in education is associated with lower support for privatized provision. Conversely, Lindh [-@lindh_public_2015] finds that in countries with more market-oriented welfare systems, support for market-based distribution tends to be higher, suggesting that individual attitudes often align with institutional outputs. These findings are consistent with neo-institutionalist and policy feedback theories [@campbell_institutional_2020], which argue that institutions do more than redistribute resources—they also shape the normative categories through which individuals assess who is deserving of support [@immergut_theoretical_1998]. In this view, dominant values and preferences are both shaped by and embedded in institutional configurations [@busemeyer_skills_2014], reinforcing the notion that institutions are not neutral structures, but active producers of the moral frameworks that legitimize or challenge inequality.

@@

The study of social mobility, its drivers and consequences has long been central to sociology. Classic theorists explored not only movement across social hierarchies but also its broader implications for class conflict, norm stability, and institutional change [@breen_effects_2024]. Sorokin [-@sorokin_social_1927] introduced the concept formally, defining mobility as the shift of individuals, values, or objects between positions within a stratification system, and distinguishing between horizontal and vertical forms. Later work differentiated intergenerational from intragenerational mobility, as well as absolute mobility—driven by structural change—from relative mobility, which captures the extent to which origins constrain destinations [@eyles_social_2022]. In Latin America, research shows high absolute but low relative mobility [@bucca_merit_2016]: although educational expansion and economic modernization have enabled some upward movement, status reproduction remains strong, especially among elites [@torche_intergenerational_2014; @lopez-roldan_comparative_2021]. This reflects deep structural inequality, segmented education systems, stratified labor markets, and legacies of dependent development that restrict access to mobility channels and reinforce the intergenerational transmission of advantage.

Beyond mapping mobility patterns, growing research has examined its subjective and attitudinal effects. Mobility effects—defined as outcomes resulting from movement between origin and destination classes [@breen_effects_2024]—have long attracted theoretical interest. Sorokin’s (1959) dissociative hypothesis posits that mobility, whether upward or downward, may produce psychological strain due to conflicting norms between class contexts, leading to lower life satisfaction. Similarly, Lenski [-@lenski_status_1954] argued that status inconsistency—mismatches among education, income, and occupation—can undermine well-being. These ideas underpin extensive empirical work linking intergenerational mobility to outcomes such as life satisfaction, mental health, and stress [@prag_subjective_2021; @gugushvili_heterogeneous_2024; @hadjar_does_2015].

Research on intergenerational social mobility has increasingly examined its effects on attitudes toward economic inequality. A key area of inquiry focuses on how upward and downward mobility influence support for redistribution, though findings remain mixed. Alesina et al. [-@alesina_intergenerational_2018] show that individuals with pessimistic expectations about their mobility—particularly those anticipating downward movement—are more likely to support generous redistributive policies. Similarly, Ares [-@ares_changing_2020] finds that upwardly mobile individuals tend to be less supportive of state-led redistribution compared to those who have experienced downward mobility. Comparative studies by Schmidt [-@schmidt_experience_2011] and Gugushvili [-@gugushvili_subjective_2017] likewise report that subjective upward mobility is associated with weaker preferences for redistribution, while downward mobility strengthens redistributive support. However, recent causal evidence from Breen and Ermisch [-@breen_effects_2024] suggests the opposite: upward mobility may increase redistributive preferences, while downward mobility may reduce them.

Beyond redistribution, other studies have explored the impact of mobility on broader beliefs about inequality. Gugushvili [-@gugushvili_intergenerational_2016] finds that upwardly mobile individuals are more likely to adopt individualistic attributions of poverty and to legitimize income inequality, particularly in post-socialist societies. Similarly, Bucca [-@bucca_merit_2016] shows that subjective upward mobility reinforces individualistic explanations of wealth in seven Latin American countries. At the macro level, Shariff et al. [-@shariff_income_2016] demonstrate that higher levels of national economic mobility correlate with greater tolerance of inequality. In contrast, Day and Fiske [-@day_movin_2017] find that low perceived mobility undermines belief in meritocracy and a just world, thereby weakening system justification. Taken together, this body of research suggests that mobility shapes attitudes toward inequality and justice through multiple, and sometimes contradictory, mechanisms.

The literature on the effects of social mobility has proposed various mechanisms to explain how and why changes in social position may influence individual outcomes [@helgason_class_2025]. One of the most prominent is the self-interest mechanism, which posits that individuals who experience upward or downward mobility undergo a shift in their material interests, thereby altering their perceptions and preferences [@helgason_longterm_2023; @ares_changing_2020; @langsaether_explaining_2022]. Closely related to this logic is the Prospect of Upward Mobility (POUM) hypothesis, which suggests that individuals may oppose redistribution not because of their current position, but because they anticipate improving their status in the future [@benabou_social_2001]. A second line of explanation draws on the framework of theories of socialization. Within this tradition, hypotheses such as acculturation, socialization, and status maximization propose that individuals adjust their attitudes based on the norms and values of either their class of origin or their destination, or a combination of both [@jaime-castillo_social_2019]. However, most of these mechanisms focus primarily on the indirect effects of origin and destination positions, rather than on the direct effect of experience of movement itself. In response to this gap, recent research has highlighted the role of self-serving bias in causal attribution processes, suggesting that individuals tend to explain their mobility trajectories in ways that justify their current position, which in turn shapes their beliefs and preferences [@schmidt_experience_2011; @gugushvili_intergenerational_2016c; @molina_its_2019].

The self-serving bias mechanism builds on a value-oriented perspective, emphasizing that individuals’ experiences of social mobility shape their causal attributions, which in turn influence their beliefs about justice and distributive preferences [@gugushvili_trends_2014]. Causal attribution refers to the process through which individuals generate explanations for their own behavior and outcomes, as well as those of others [@gugushvili_intergenerational_2016]. In this view, people’s interpretations of economic inequality depend on whether they believe such disparities reflect unequal individual contributions. Individuals who adopt an internal attribution framework tend to see success or failure as rooted in personal characteristics such as effort, talent, or merit. In contrast, those who regard inequality as unjust are more likely to adopt an external attribution model, viewing outcomes as the result of structural barriers beyond individual control [@kluegel_beliefs_1981]. This mechanism, often described as intrapersonal causal attribution, reflects how people explain their own socioeconomic positions—typically attributing their successes to internal qualities while blaming failures on external circumstances [@miller_selfserving_1975]. Over time, individuals may revise their beliefs and attitudes: while early views are shaped by their social origin, these are later adjusted in light of personal experiences of mobility and the perceived role of ascribed versus achieved factors in determining socioeconomic outcomes [@gugushvili_intergenerational_2016].

Empirical research has provided support for the self-serving bias mechanism in shaping redistributive preferences and attitudes toward the legitimacy of inequality. For instance, Schmidt [-@schmidt_experience_2011] finds that individuals who experience upward mobility are more likely to interpret their success as the result of personal effort or merit, and consequently perceive redistribution as less necessary. Conversely, individuals who experience downward mobility tend to attribute their decline to external circumstances—such as structural inequality or unemployment—and show stronger support for redistribution. In a comparative analysis across different welfare domains, Gugushvili [-@gugushvili_subjective_2017] demonstrates that upward mobility is associated with lower support for government spending on housing and pensions, while individuals who experience downward mobility express lower support for healthcare and education spending, but favor increased investment in housing and pensions, reflecting the material nature of these domains. Moreover, Gugushvili [-@gugushvili_intergenerational_2016c] finds that upward mobility is linked to greater justification of income inequality, suggesting that improved social standing reinforces an attributional view in which success is seen as the result of individual characteristics—thus legitimizing inequality as a fair outcome.

Consequently, considering this theoretical and empirical background, the first hypothesis of this research is that:

$H1$: Experiencing upward (downward) social mobility is positively (negatively) associated with greater support for market justice in healthcare, pensions, and education.

@@

Meritocracy constitutes a central ideological framework for legitimizing different types of social inequality, for instance through market justice beliefs. Rooted in the belief that rewards and positions should be allocated based on individual effort and talent, meritocracy operates as a normative ideal and a descriptive belief about how society functions. As initially conceptualized by Michael Young [-@young_rise_1958], the term was meant to critique a system in which merit-based stratification becomes a new form of inequality. However, over time, meritocracy has been widely supported in many societies as a fair and desirable principle of distribution, particularly within liberal democracies and market-oriented economies [@mijs_paradox_2019; @sandel_tyranny_2020]. From a sociological perspective, the belief in meritocracy is more than a cognitive assessment; it reflects a moral lens through which individuals interpret inequality. People who believe that success results from hard work and talent are more likely to view social and economic disparities as legitimate [@castillo_meritocracia_2019; @batruch_belief_2023]. Conversely, if they see outcomes as driven by luck, social origin, or systemic barriers, inequality is more likely to be perceived as unjust. This distinction becomes crucial in societies with persistent structural inequality, where public narratives often emphasize personal responsibility and merit while overlooking entrenched disadvantages.

I adopt a multidimensional perspective on meritocracy, distinguishing between two key dimensions: effort-based and talent-based perceptions. This distinction is essential, as it captures different pathways through which individuals justify inequality [@young_rise_1958]. Effort-based meritocracy emphasizes hard work and perseverance as the basis for social rewards, aligning closely with cultural narratives of personal responsibility. A talent-based meritocracy, by contrast, emphasizes intelligence and innate abilities, which are often perceived as less malleable and more unequally distributed. Both dimensions have been shown to correlate with acceptance of inequality, but they may carry distinct implications for how inequality is justified in specific domains [@castillo_multidimensional_2023]. The relevance of this distinction is supported by recent studies, which show that individuals respond differently to these dimensions. For instance, perceptions that effort is rewarded in society are more strongly associated with positive evaluations of fairness and acceptance of unequal outcomes [@batruch_belief_2023; @wiederkehr_belief_2015; @wilson_role_2003]. This may be because effort is seen as a controllable and morally virtuous trait, whereas talent is often perceived as a natural advantage. Consequently, effort-based meritocracy is likely more potent in legitimizing inequality, particularly in neoliberal contexts.

These dimensions of meritocracy reflect how respondents perceive society's distributive logic, regardless of whether they endorse meritocratic principles. This distinction aligns with recent findings indicating that individuals distinguish between how merit is perceived in society and how it should ideally operate, which in turn shapes their preferences for redistribution and justice [@tejero-peregrina_perceived_2025]. Meritocratic beliefs serve as symbolic justifications for unequal outcomes, particularly when access is stratified by income or social opportunity. Prior studies in Chile have shown that individuals who perceive higher levels of meritocracy tend to express stronger support for unequal distributions that reflect market outcomes in healthcare, education and pensions [@castillo_perceptions_2025; @castillo_socialization_2024].

In addition to influencing individual attitudes toward inequality, meritocratic beliefs can contribute to social division and the stigmatization of disadvantaged groups. Recent research has demonstrated that exposure to meritocratic narratives can reinforce the belief that poverty results from individual failings rather than systemic conditions, reducing support for redistributive measures and increasing the stigmatization of the poor [@hoyt_mindsets_2023]. This reinforces negative stereotypes and reduces empathy toward individuals from lower socioeconomic backgrounds. Moreover, Busemeyer et al. [-@busemeyer_positive_2021] argues that meritocratic narratives can serve as feedback mechanisms that shape public opinion and well-being by framing individuals' understanding of welfare outcomes as deserved or undeserved within existing institutional structures. This psychological mechanism highlights the normative power of meritocracy in stabilizing unequal systems by shaping political attitudes and personal perceptions of success and failure.

>>>>

---
bibliography: "../input/bib/tesis-magister.bib"
csl: "../input/bib/apa6.csl"
---

@@

## Preferences for pensions commodification

Beyond the state’s capacity to redistribute resources, market institutions are central arenas for allocating socially valuable goods and risks [@lindh_bringing_2023]. Markets are not mere aggregates of individual choices but social institutions endowed with rules and normative meanings, built through “moral projects, saturated with normativity” that permeate everyday thinking around fairness, effort and deservingness [@fourcade_moral_2007; @koos_moral_2019]. In this sense, the economic order is mirrored in a moral economy: shared norms and beliefs about what is considered a just distribution, embedded and reinforced through institutions and policy feedback [@svallforsMoralEconomyClass2006a]. In an era of privatisation and commodification, market logic has expanded into core areas of social reproduction—childcare, healthcare, education and pensions—stratifying access and quality [@ferre_welfare_2023; @gingrich_making_2011] and generating feedback effects on public beliefs about welfare and market inequality [@lindh_public_2015]. In this context, support for market-based welfare provision has grown worldwide, particularly among higher-income groups who view private alternatives as more efficient or higher-quality [@busemeyer_welfare_2020; @lindh_public_2015]. These institutional transformations provide the backdrop for analysing how far citizens accept the commodification of social protection [@satz_por_2019].

The legitimacy of market-based welfare is closely tied to beliefs about distributive justice grounded in market principles. Beyond stratification structures, citizens hold structured beliefs about how resources ought to be allocated, reflecting causal attributions for inequality, normative principles and expectations about deservingness [@kluegel_beliefs_1981]. Research on distributive justice examines conceptions of how goods and rewards should be distributed and when inequality is considered just [@castillo_legitimacy_2011; @jasso_gender_1999], while work on redistributive preferences focuses on support for state-led mechanisms that reduce inequality [@cavaille_fair_2025]. Situated at this intersection, the emerging literature on market justice preferences conceptualises the market itself as a site of redistribution and asks whether individuals regard market criteria as fair bases for allocating wages and access to core welfare services [@lindh_public_2015; @lindh_bringing_2023; @castillo_perceptions_2025]. In this perspective, the expansion of market logic into welfare domains implies that social services are treated as legitimate commodities that can be priced and stratified [@busemeyer_welfare_2020], and empirical work suggests that such institutional changes feed back into attitudes, reinforcing market-conforming understandings of fairness and responsibility [@zhu_policy_2015].

Market justice preferences refer to normative beliefs that legitimise allocating goods and services according to market criteria such as income and purchasing power [@kluegel_legitimation_1999; @lindh_public_2015]. Building on Lane’s [-@lane_market_1986] contrast between market and political justice, market justice is a distributive principle that endorses rewarding effort, productivity and skill rather than need and equality as emphasised in welfare-state policies. In this study, I treat market justice less as an abstract doctrine and more concretely as preferences for commodification: support for treating welfare services as commodities, so that better services follow from higher income or ability to pay, and the resulting stratified access is seen as fair [@lindh_public_2015]. Empirically, such orientations are typically measured with items asking whether it is fair that access to services depends on income, a strategy rooted in research on justifications of capitalist inequality [@kluegel_legitimation_1999] and applied to healthcare and education [@lee_fairness_2023; @vondemknesebeck_are_2016; @immergut_it_2020; @castillo_changes_2025]. Moving beyond single domains, @lindh_public_2015 develops a comparative index of income-based access to healthcare and education, while @castillo_perceptions_2025 use the same single-item measure but add the pensions domain in Chile. These constructs gauge how individuals view market-generated inequalities as legitimate and capture two core dimensions of market distribution: the role of income in determining attainment and the framing of services as commodities that can be bought and sold according to ability to pay [@lindh_public_2015].

This study examines support for the commodification of pensions, defined as normative judgements on whether better pensions should depend on income or individual contributions. I treat this as a domain-specific form of market justice: pensions are conceived as commodities whose allocation should follow contribution- and return-based criteria rather than principles of need [@lane_market_1986; @mau_inequality_2015]. The spread of funded, privately managed schemes built on individual accounts has transformed both the distributive and normative structure of old-age security [@huber_political_2000; @arenas_sistemas_2019]. Distributively, these schemes reassign risk from the collective to the individual by tying benefits to formal employment, contribution density, earnings and financial returns, making retirement income heavily dependent on labour-market trajectories and capital-market success [@madero-cabib_private_2019; @oecd_pensions_2023]. Normatively, they recast pensions as the outcome of personal investment decisions rather than collective obligations, promoting ideals of individual responsibility, self-economisation and contribution-based desert [@arza_pension_2008; @borzutzky_pension_2012]. Contributions are increasingly framed as private assets rather than risk pooling, and workers are expected to act as employee–entrepreneurs who choose providers, monitor portfolios and “make their money work” over the life course [@kerner_pension_2020; @mau_inequality_2015; @borzutzky_chiles_2016]. As more people tie their security expectations to private investments and financial markets, their material interests become more deeply rooted in the market, and their imperatives gain stronger normative support [@mau_inequality_2015]. When citizens are asked whether higher pensions should be allocated to those who earn and contribute more, they reflect support for the market allocation of a critical social service and the legitimacy of the resulting risks and inequalities in old age [@busemeyer_welfare_2020; @lindh_public_2015].

@@

The drivers and consequences of social mobility have long been central to sociology. Social mobility denotes movements between positions in a stratification system. Classical and later work distinguish intragenerational (within the life course) from intergenerational mobility (between parents and children), as well as absolute mobility—driven by structural change—from relative mobility, which captures the extent to which social origins constrain destinations [@eyles_social_2022]. Beyond mapping mobility patterns, a large body of research examines how mobility shapes attitudes and behaviours. In this literature, “mobility effects” refer to outcomes that arise from movement between origin and destination classes [@breen_effects_2024]. A growing number of studies analyse the consequences of intergenerational mobility for attitudes towards economic inequality [e.g. @alesina_intergenerational_2018; @bucca_merit_2016; @day_movin_2017; @gugushvili_intergenerational_2016; @helgason_class_2025]. These inquiries rest on the idea that, if attitudes to inequality are strongly stratified by class—because classes embody distinct material interests and moral economies [@edlundDemocraticClassStruggle2015a; @kulinClassValuesAttitudes2013]—, then people who move between classes should adjust their attitudes towards those typical of their new class location. This study contributes to this line of work by examining how intergenerational mobility is associated with preferences for the commodification of pensions in a highly stratified and commodified welfare regime such as Chile’s.

The literature on the effects of social mobility identifies several mechanisms through which changes in class position may affect individual attitudes [@helgason_longterm_2023]. A first family of accounts emphasises material self-interest and can be divided into myopic and anticipatory theories. Myopic theories assume that individuals rapidly align their attitudes with their current class position, so mobility triggers shifts in material interests, information and perceived risks that update preferences [@helgason_longterm_2023; @ares_changing_2020; @langsaether_explaining_2022]. Anticipatory theories stress that forward-looking agents align their current attitudes with expected future income, as in the Prospect of Upward Mobility (POUM) hypothesis, in which individuals may oppose redistribution because they anticipate moving up [@benabou_social_2001].

A second family foregrounds culture and socialisation. Acculturation accounts posit that people gradually adapt their views to those prevailing in their destination class through exposure to new information, class-homophilous networks, group pressure and the assimilation of principles and normative beliefs [@jaime-castillo_social_2019; @helgason_class_2025]. Socialisation perspectives, instead, argue that core attitudes are formed early in life in the class of origin through families, schools, and critical events, and remain largely stable despite later mobility [@jaime-castillo_social_2019]. Status-maximisation theories propose that individuals align their attitudes with the highest class they have occupied, so upwardly mobile respondents converge towards their destination class, while downwardly mobile respondents retain attitudes from their higher-status origins [@jaime-castillo_social_2019]. Helgason and Rehm [-@helgason_longterm_2023;-@helgason_class_2025] refine these ideas in a learning framework, in which attitudes evolve slowly through cumulative class experiences that combine economic resources, socialisation into class-specific values, information, and networks. Finally, attributional mechanisms highlight self-serving biases, suggesting that individuals explain their mobility trajectories in ways that legitimise their current position, thereby shaping beliefs about effort, deservingness and inequality [@gugushvili_intergenerational_2016; @molina_its_2019; @schmidt_experience_2011]. In what follows, I focus on two mechanisms that are particularly relevant for market-based pensions: material self-interest and acculturation into meritocratic distributive norms.

How intergenerational social mobility shapes preferences for the commodification of pensions? A plausible mechanism runs through changes in material self-interest. Social mobility alters individuals’ incomes, job security and access to economic resources [@helgason_longterm_2023; @mau_inequality_2015], thereby modifying the costs and benefits they face in market-based pension systems. Material interests can be understood as the pursuit of economic well-being under the constraints and trade-offs implied by one’s class position [@wrightClassCountsComparative1997a]. Upward mobility typically entails higher and more stable earnings, greater capacity to contribute to funded schemes, greater tolerance for investment risk, and the ability to purchase supplementary coverage, whereas downward mobility reduces income, undermines contribution histories, and increases dependence on public or solidarity-based pillars [@madero-cabib_private_2019; @galvez_pensiones_2024]. These diverging pension prospects imply that upwardly and downwardly mobile individuals face systematically different stakes in market-based pension systems, making material self-interest a first channel through which mobility may shape support for pension commodification.

Consistent with this reasoning, existing research indicates that support for welfare commodification and pension marketisation is strongly structured by material advantage. Individuals with higher income, education and occupational class are more likely to endorse market-based distributive principles and to view private provision as a way to secure or enhance relative advantage  [@koos_moral_2019; @vondemknesebeck_are_2016; @immergut_it_2020; @busemeyer_skills_2014; @lindh_public_2015; @svallfors_political_2007]. In the pension field, institutional design sharpens these incentives. @busemeyer_welfare_2020 show that in countries where private alternatives to public pensions exist, support for public pension spending among upper-income groups erodes as they exit public schemes into private plans whose contributions and benefits closely track their income and risk. In Latin America, @kerner_pension_2020 finds that higher income and higher perceived financial returns from pension funds are associated with preferences for a stronger market role in pensions, suggesting that favourable returns are interpreted as evidence that pension neoliberalism “works”. Evidence from Chile points in the same direction: higher-income and university-educated respondents express stronger support for market allocation in healthcare, education and pensions [@castillo_socialization_2024; @otero_power_2024], and economic elites—especially those socialised in highly prestigious schools—hold more favourable attitudes towards inequality that can be traced to their material interests against redistribution [@carranza_what_2024]. Systems that link benefits to individual contributions thus allow better-off groups to capitalise on their financial independence and minimise reliance on public support, implying that mobility into or out of these advantaged positions should alter individuals’ stakes in pension commodification.

Research on redistribution and inequality offers additional, indirect support for this self-interest mechanism. Individuals with pessimistic mobility expectations—especially those anticipating downward movement—display stronger support for redistributive policies [@alesina_intergenerational_2018]. Studies tracking realised mobility show that, compared to immobility, upward moves are associated with more economically conservative, less redistributive attitudes, whereas downward moves strengthen demand for state intervention and redistribution [@ares_changing_2020; @langsaether_explaining_2022; @schmidt_experience_2011; @jaime-castillo_social_2019; @rodriguez_intergenerational_2024]. Over the life course, cumulative exposure to higher-class positions—whether by staying or moving up—is linked to more market-liberal economic values, while sustained exposure to lower-class positions—through stability or downward mobility—reinforces pro-redistributive orientations [@helgason_longterm_2023;@helgason_class_2025]. Domain-specific evidence also aligns with changing material stakes: @gugushvili_subjective_2017 finds that upward mobility is associated with lower support for government spending on housing and pensions, whereas downward mobility reduces support for healthcare and education but increases support for housing and pensions, reflecting the material nature of these domains. Beyond redistribution, upward mobility fosters more individualistic attributions of poverty and wealth and greater legitimation of income inequality [@bucca_merit_2016; @gugushvili_intergenerational_2016; @gugushvili_intergenerational_2016c], and societies with higher economic mobility display more tolerance of inequality [@shariff_income_2016]. Taken together, this evidence suggests that social mobility realigns attitudes with changing material stakes and distributive interests. Although the present study cannot directly test the underlying mechanism, it treats myopic self-interest as the central interpretive framework for the expected effects of intergenerational mobility on support for pension commodification.

Consequently, the main hypothesis of this research is:

$H_1$: Compared to immobility, intergenerational upward mobility is associated with stronger, and downward mobility with weaker, support for the commodification of pensions.

While material self-interest is the main lens, mobility also reshapes norms and interpretive frames. Mobile individuals move across social milieus and networks and are exposed to new standards and beliefs about fairness and deservingness [@helgason_longterm_2023]. Acculturation accounts hold that people gradually adopt the attitudes of their destination class [@jaime-castillo_social_2019]. Mobility can thus reshape how individuals interpret success, failure and inequality, with upward moves fostering more individualistic and meritocratic understandings and downward moves encouraging structural accounts of disadvantage [@gugushvili_subjective_2017; @mau_inequality_2015; @mijs_belief_2022] Applied to pensions, those who rise may come to see market-based schemes as fair systems that reward effort and foresight. In contrast, those who fall may view commodified pensions as unjust and insufficiently protective. Therefore, shifts in meritocratic norms can be a second channel through which the mobility effect operates, serving as a complementary acculturation mechanism that may influence support for pension commodification, as examined in the next section.

@@

Meritocracy refers to a distributive system in which individual merit—typically defined as effort and talent—is the primary criterion for allocating resources and rewards, rather than social origins or inherited privilege [@young_rise_1958; @sen_merit_2000]. Originally coined by Young in a dystopian critique of a society where power and status are justified by achievement and mobility only driven by “merit”, the term has been re-appropriated as a positive ideal of fairness, especially in liberal and market-oriented societies [@dubet_repensar_2011; @mijs_paradox_2019; @vandewerfhorst_meritocracy_2024]. From a sociological perspective, belief in meritocracy is not just a cognitive judgement; it constitutes a moral lens through which individuals interpret social and economic disparities [@castillo_multidimensional_2023]. The conviction that economic inequalities are justified because they reflect differential merit has been identified as a key mechanism behind the persistence of inequality, since it recasts structural advantages and disadvantages as outcomes of individual performance and encourages “winners” to see their position as earned and “losers” to internalise blame rather than question the underlying structure of opportunity [@mijs_unfulfillable_2016; @sandel_tyranny_2020; @wilson_role_2003]. Understanding how these beliefs in meritocracy are structured is therefore crucial for analysing attitudes toward inequality.

Recent research has emphasised the need to decompose the term “meritocratic beliefs” and to distinguish between meritocratic preferences and meritocratic perceptions [@castillo_multidimensional_2023; @zhu_meritocratic_2025]. Preferences refer to normative ideas about how rewards should be allocated—whether people think effort and ability ought to determine life chances—whereas perceptions capture evaluations of how meritocracy actually operates in society, that is, whether people believe that existing inequalities reflect merit-based processes [@janmaat_subjective_2013]. This distinction is crucial for understanding how people interpret inequality in stratified societies because it clarifies the (mis)alignment between normative support for meritocracy and evaluations about its actual implementation [@castillo_multidimensional_2023; @mijs_paradox_2019; @mijs_how_2022; @newman_economic_2023; @lindner_does_2024]. In this study, I focus on meritocratic perceptions of effort and talent as key dimensions of broader meritocratic beliefs, given their central role in shaping attitudes toward inequality and welfare.

Several empirical studies have examined the social foundations and consequences of meritocratic perceptions. Individuals with higher levels of education, income and occupational prestige are more likely to endorse merit-based explanations for social outcomes [@mijs_paradox_2019; @duru-bellat_who_2012; @garcia-sanchez_perceptions_2018]. These perceptions function as a normative framework that legitimises several unequal outcomes, especially when access to social goods is stratified by income or social background. Individuals who see their society as meritocratic tend to show lower support for redistribution [@hoyt_mindsets_2023; @tejero-peregrina_perceived_2025], greater legitimisation of class inequality [@darnon_where_2018], increased tolerance of income gaps [@batruch_belief_2023], more support for system-justifying ideologies [@wiederkehr_belief_2015], and even lower perceived levels of economic inequality itself [@castillo_meritocracia_2019]. This evidence provides a basis for extending the study of the attitudinal consequences of meritocracy to preferences for market-based welfare.

Preferences for the commodification of social welfare services are anchored in normative frames. Studies show that support for market-based welfare is higher among individuals who endorse meritocratic and liberal principles—that benefits should depend on individual effort and contributions—and among those with economically conservative orientations, whereas egalitarian views reduce support for market allocation [@lee_fairness_2023; @quadagno_has_2012; @jaime-castillo_public_2013]. In the pension field, factorial evidence from Chile indicates that merit-based criteria such as educational attainment and years in the labour force weigh more heavily than need-based attributes in judgments of just pension levels, with respondents willing to accept very low pensions for those perceived as low achievers [@castillo_deserving_2019]. Recently, @castillo_perceptions_2025 show that agreement with the notion that it is fair for higher-income individuals to access better welfare services was relatively low in 2016 but rose significantly by 2023, especially in the pension domain. They also find that individuals who endorse more meritocratic perceptions–specifically the perception that effort is rewarded– are more likely to support market-based access for healthcare, education and pensions. All in all, meritocratic perceptions are a powerful source of inequality legitimation, anchoring who is seen as deserving to be better or worse off, and are strongly linked to how individuals understand unequal life chances and movements up and down the social hierarchy.
